---
title: "Homework 5 Bivariate Relationships"
author: "Walid Medani"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_handout: 
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
always_allow_html: true
---

```{css}
.chunkcolor {
background-color: MistyRose;
}
options(scipen = 9999)
```

```{r setup, include=FALSE}
library(tufte)
library(tidyverse)
library(tinytex)
library(kableExtra)
library(ggthemes)
library(knitr)
library(ggplot2)
library(dplyr)
library(gapminder)
library(ggdark)
```

# Introduction

Due to my previous dataset being limited in showcasing bi-variate relationships, I've chosen the gapminder dataset that explores population, life expectancy, and GDP of all countries from 1952 to 2007.

Below are the summary statistics of the dataset:

```{r intro, class.source="chunkcolor"}
summary(gapminder) %>%
  kbl() %>%
  kable_material_dark(c("striped", "hover"))
  
```

Countries Cuba and South Africa are randomly selected to observe life expectancy.

```{r fsf, class.source="chunkcolor"}
gapminder %>% 
  select(country, lifeExp) %>% 
  filter(country == "Cuba" |
         country == "South Africa") %>% 
  group_by(country) %>% 
  summarise(avglife = mean(lifeExp)) %>% 
  kbl() %>% 
  kable_material_dark(c("striped", "hover"))
```

Is the difference between Cuba and South Africa's life expectancy due to chance? Since I'm comparing the means of two groups, a t-test can validate the difference in life expectancy and conclude that it isn't due to chance in the sample.

```{r ttest, class.source="chunkcolor"}
southcuba <- gapminder %>% 
  select(country, lifeExp) %>% 
  filter(country == "Cuba" |
         country == "South Africa")

t.test(data = southcuba, lifeExp ~ country)
```

# Linear Regression

Is there a correlation between life expectancy and gdp per capital and population as seen in the graph below? To find out, I use a multivariate linear regression model by placing GDP & population as the explanatory variables. With p-values below 0.05, our model concludes that there is in fact a correlation between life expectancy and gdp + population.

```{r graph, message=FALSE, warning=FALSE, class.source="chunkcolor"}
gapminder %>%
  filter(gdpPercap < 50000) %>% 
  ggplot(aes(
          x =log(gdpPercap), 
          y=lifeExp,
          color=continent,
          size=pop)) +
  geom_point(alpha=0.2)+
  geom_smooth(method = lm)+
  labs(title = "Correlation between GDP per capita & life expectancy",
       subtitle = "1952 - 2007") +
  theme_fivethirtyeight()+
  ggdark::dark_mode()+
  facet_wrap(~continent)
```

```{r linear regression, class.source="chunkcolor"}
summary(lm(lifeExp ~ gdpPercap+pop, gapminder))
```
